# Analysis of Contributions towards Presidential Campaign in 2016 in California
 by Rajat Gupta
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}

library(ggplot2)
library(ggmap)
library(mapproj)
library(plyr)
```

```{r echo=FALSE, Load_the_Data}
# Load the Data
Data <- read.csv('P00000001-CA.csv', header = TRUE, sep = ",", 
                 row.names = NULL)

#As all the columns are shifted by one due to the appearance of 'row.names' 
#because of the way the data has been imported, and the last coulumn is empty, 
#the data will be shifted to the right and then, this extra column will 
#be deleted. 
Data[2:19]<-Data[1:18]
Data <- subset(Data, select = -c(row.names))

clean <- function(original, vals_to_change, final_val){
  final <- ifelse(original %in% vals_to_change,
                  final_val, as.character(original))
  return (factor(final))
}

dem <- c("Clinton, Hillary Rodham",
         "Sanders, Bernard",
         "O'Malley, Martin Joseph",
         "Webb, James Henry Jr.",
         "Lessig, Lawrence")

oth <- c("Johnson, Gary",
         "McMullin, Evan",
         "Stein, Jill")

fem <- c("Clinton, Hillary Rodham",
         "Stein, Jill",
         "Fiorina, Carly")

self <- c("SELF-EMPLOYED","SELF", "SELF EMPLOYED-EMPLOYED")

Data$pol_party <- ifelse(Data$cand_nm %in% dem, 'Democratic Party', 
                         ifelse((Data$cand_nm %in% oth),
                                'Other', 'Republican Party'))
Data$pol_party <- factor(Data$pol_party)

Data$gender <- ifelse(Data$cand_nm %in% fem, 'Female', 'Male')
Data$gender <- factor(Data$gender)


#Data in the column contbr_employer contains mainy abreviations and duplicates,
#thus it needs to be cleaned which is being done here. Due to the volume of 
#entries, changes are made for values having large entries only.
Data$contbr_employer <- clean(Data$contbr_employer, self, "SELF EMPLOYED")
Data$contbr_employer <- clean(Data$contbr_employer, c("APPLE INC."), "APPLE")
Data$contbr_employer <- clean(Data$contbr_employer, c("ORACLE CORP."), 
                              "ORACLE")
Data$contbr_employer <- clean(Data$contbr_employer, c("UNEMPLOYED"), 
                              "NOT EMPLOYED")
Data$contbr_employer <- clean(Data$contbr_employer, c("UCSD"), 
                              "UNIVERSITY OF CALIFORNIA SAN DIEGO")
Data$contbr_employer <- clean(Data$contbr_employer, c("USC"), 
                              "UNIVERSITY OF SOUTHERN CALIFORNIA")
Data$contbr_employer <- clean(Data$contbr_employer, c("SFUSD"), 
                              "SAN FRANCISCO UNIFIED SCHOOL DISTRICT")
Data$contbr_employer <- clean(Data$contbr_employer, c("UC BERKELEY"), 
                              "UNIVERSITY OF CALIFORNIA, BERKELEY")
Data$contbr_employer <- clean(Data$contbr_employer, c("KAISER"), 
                              "KAISER PERMANENTE")
Data$contbr_employer <- clean(Data$contbr_employer, c("STATE OF CA"), 
                              "STATE OF CALIFORNIA")
Data$contbr_employer <- clean(Data$contbr_employer, c("UCSF"), 
                              "UNIVERSITY OF CALIFORNIA SAN FRANCISCO")
Data$contbr_employer <- clean(Data$contbr_employer, 
                              c("INFORMATION REQUESTED PER BEST EFFORTS"), 
                              "INFORMATION REQUESTED")
#making the empty values to be N/A
Data$contbr_employer <- clean(Data$contbr_employer, c(""), "N/A")

```

This dataset looks at the Financial Contributions made to the presidential 
campaign for the U.S. presidential election in 2016 in the State of California.
The data is provided by the Federal Election Commission and includes 
information about the donor, the amount contributed and the receiving 
candidate. Extra information about the individual's party affiliation and 
gender have also been added.

# Univariate Plots Section

Looking at the Data, the variables within it,

```{r echo=FALSE, message=FALSE, warning=FALSE}
dim(Data)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
str(Data)
```


```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(Data)
```

We see that most of the data contains ordered factors, with one continuos 
variable which will be studied extensively ('contb_receipt_amt').

Now, looking at variation in different variables,

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(contb_receipt_amt), data= Data)+
  geom_histogram(binwidth = 10)+
  scale_x_continuous(breaks = seq(0,1000,50), limits = c(0,1000) )+
  ylim(0, 300000)+
  xlab('Contribution Amount')
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(subset(Data, contb_receipt_amt>0)$contb_receipt_amt)
```

We can clearly see that most of the donations made are small amounts, with very 
few large donations.

```{r echo=FALSE, message=FALSE, warning=FALSE}
large_contrib_cities <- (summary(Data$contbr_city)>15000)
large_contrib_cities <- names(large_contrib_cities[large_contrib_cities 
                                                   == TRUE])
ggplot(aes(contbr_city), 
       data = subset(Data, contbr_city %in% large_contrib_cities))+
  geom_bar()+xlab('City Name')+
  theme(axis.text.x = element_text(angle=90, hjust=1)) 
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
city_vals <- ddply(Data, .(contbr_city), "nrow")
city_vals <- subset(city_vals, nrow>35000)
city_vals <- droplevels(city_vals)

xx=c("lon", "lat")
city_vals[xx] <-geocode(as.character(city_vals$contbr_city))

Cal <- subset(map_data("state"), region %in% c("california"))

Cal_map <- ggplot(data=Cal)+
  geom_polygon(aes(x=long,y=lat), color='black', fill='yellow')+
  coord_fixed(1.3)

Cal_map+
  geom_point(aes(x=lon, y=lat, size = nrow), data=city_vals, alpha=0.2)+
  scale_size_area(name="No. of Donations")+
  geom_text(aes(x=lon, y=lat, label=contbr_city), hjust="inward", 
            data=subset(city_vals, nrow>35000))
```


```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(Data$contbr_city)[summary(Data$contbr_city)>15000]
large_contrib_cities <- names(summary(Data$contbr_city)
                              [summary(Data$contbr_city)>15000])
```

We can see that the largest number of the contributions are coming from cities 
located on the California coast (Los Angeles, San Francisco & San Diego).

```{r echo=FALSE, message=FALSE, warning=FALSE}
large_contrib_cand <- (summary(Data$cand_nm)>10000)
large_contrib_cand <- names(large_contrib_cand[large_contrib_cand == TRUE])
ggplot(aes(cand_nm), 
       data = subset(Data, cand_nm %in% large_contrib_cand))+
  geom_bar(width = 0.7)+
  xlab('Candidate Name')+
  theme(axis.text.x = element_text(angle=90, hjust=1)) 
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot( aes(pol_party), data=Data)+
  geom_bar(width = 0.5)+
  ylab('Number of Contributions')+
  xlab('Political Party')
```

As the candidates who got the most donations were affiliated with the 
Democratic Party, it is no surprise that their party had the largest number of 
contributions in the state of California.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(gender), data = Data)+
  geom_bar(width = 0.5)
```

We can see that Female candidates received greater number of donations than 
their male counterparts.

Lets, look at the employers for the donating individuals.

```{r echo=FALSE, message=FALSE, warning=FALSE}
print("Donor Occupation information")
head(summary(Data$contbr_occupation))
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
top_employers <- names(summary(Data$contbr_employer)
                       [summary(Data$contbr_employer)>1500])
ggplot(aes(contbr_employer), data=subset(Data, 
                                         contbr_employer %in% top_employers))+
  geom_bar()+
  scale_y_log10()+
  xlab('Employer Name')+
  coord_flip()
  
```

We can see that most of the individuals that contributed financially are either
Self Employed, retired or the data is unavailable. The above graph has been 
created after extensive data cleaning because of abbreviations and empty cell 
values which have been renamed N/A due to lack of knowledge.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(election_tp), data = Data)+
  geom_bar(width = 0.7) +
  scale_y_log10()+
  xlab('Election Stage')
```

While analysing, the stages at which the donations were made, we see that the 
column "election_tp" has 1695 empty entries. There are also 10548 empty entries
in the column, "contbr_employer".

```{r echo=FALSE, message=FALSE, warning=FALSE}
top_dates <- (summary(Data$contb_receipt_dt)>9000)
top_dates <- names(top_dates[top_dates == TRUE])
ggplot(aes(contb_receipt_dt), 
       data=subset(Data,contb_receipt_dt %in% top_dates))+
  geom_bar()+
  xlab("Top Donation Dates")+
  theme(axis.text.x = element_text(angle=90, hjust=1)) 
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
top_dates
```

We find that most of the donations were either made in the months of October 
and Novemeber, close to the General Election in the country or during the 
months of March and April, before the Primary Election in Califronia.

# Univariate Analysis

### What is the structure of your dataset?
The data has been divided into various columns providing information about the 
donor, the candidate to whom the donation has been made and when the donation
was made. This information has been spread across 20 variables of which 17
variables are ordered factors, only one is a continous variable 
(contb_receipt_amt).

### What is/are the main feature(s) of interest in your dataset?
The main features of interest involve looking at the relation between the 
amount donated, number of contibutions and the characterstics (occupation, 
employer) of the donor. It will also be interesting looking at different 
candidates and analysing the sources of their donations.

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?
The information about party affiliation, candidate gender, contributed amount,
time at which the contribution was made, all lend themselves to the analysis
being performed and will help in determining if any correlation exists between
different variables.

### Did you create any new variables from existing variables in the dataset?
Two new variables describing the party affiliation and the gender of the 
candidate were added to help in analysing the data. Data for the creation of 
both the variables was achieved through the Wikipedia page for each of the 
candidates.

This data is required to analyse the variation in donations made according to
the party affiliation and candidate gender to determine possible donor 
preferences.

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

The data has a lot of empty entries specially in the election_tp, 
contbr_employer variables. 

The variable, contbr_employer was cleaned to care of over abbreviated employer
names and empty values. However, 'complete' cleaning was not possible due to 
high number of records and presence of unique entries each following a 
different standard.

Also, while analysing the amount contributed, only values with positive 
contb_receipt_amt were considered because a negative value represented a refund
made back to the donor which doesn't reflect a donation.

# Bivariate Plots Section

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(contb_receipt_amt), data = subset(Data, contb_receipt_amt>0))+
  geom_histogram( binwidth = 10)+
  scale_x_continuous(limits = c(0,1010), breaks =seq(0,1010,100) )+
  facet_wrap(~gender) +
  xlab("Donation Amount")
```


```{r echo=FALSE, message=FALSE, warning=FALSE}
by(subset(Data,contb_receipt_amt>0)$contb_receipt_amt,
   subset(Data,contb_receipt_amt>0)$gender, 
   summary)
```

It is interesting to note that although, the maximum & median donation made to
male candidates was greater than that for female candidates, the mean donation 
made to the female candidates was higher, this could be due to a large number 
of high donations made to female candidates compared to their male 
counterparts. Let's verify this by looking at donations greater that $1000.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(contb_receipt_amt), data = subset(Data, contb_receipt_amt>1010))+
  geom_histogram( binwidth = 50)+
  facet_wrap(~gender) +
  xlim(1010,3000)+
  xlab("Donation Amount")
```

It was interesting to note that there appeared to be a spike in donations made 
near $3000 for both male and female candidates, but this spike is greater for 
female candidates than for male candidates, which would explain the higher mean
value for female candidates.

Let's look at which Female candidates is receiving these donations,

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(contb_receipt_amt), data = subset(Data, (contb_receipt_amt>1010)&
                                               (gender == 'Female')))+
  geom_histogram( binwidth = 50)+
  facet_wrap(~cand_nm) +
  xlim(1010,3000)+
  xlab("Donation Amount")

```

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(contb_receipt_amt), data = subset(Data, (contb_receipt_amt>1010)&
                                               (gender == 'Female')))+
  geom_histogram( binwidth = 50)+
  facet_wrap(~cand_nm) +
  xlim(3000,6000)+
  xlab("Donation Amount")
```

It is interesting to see from these graphs that "Clinton, Hillary Rodham" 
recieved much greater donations between 2500 and 3000 than rest of the female 
candidates. But, "Fiorina, Carly" received greater number of donations ranging 
from 5000-6000$. As we see that the number of donations received by "Clinton, 
Hillary Rodham" were much greater than others, she received the most donations 
among female candidates.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=pol_party, y=contb_receipt_amt), 
       data = subset(Data, contb_receipt_amt>0))+
  geom_boxplot()+ 
  coord_cartesian(ylim=c(0,1000))+
  xlab("Political Party")+
  ylab("Amount Contributed")
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
by(subset(Data,contb_receipt_amt>0)$contb_receipt_amt, 
   subset(Data,contb_receipt_amt>0)$pol_party, 
   summary)
```

This verifies the boxplots achieved, telling us that although the Democratic 
Party received greater number of donations (from Univariate Analaysis), the 
Republican and the other parties received a higher mean and median contribution
amount per donor.

Lets look at the Individuals who have made the highest single contributions to
the campaigns,

```{r echo=FALSE, message=FALSE, warning=FALSE}
high_payers <- (subset(Data,contb_receipt_amt>=5500))
print("Occupation of Top Paying Individuals:")
summary(high_payers$contbr_occupation)[summary
                                       (high_payers$contbr_occupation)>0]
```

Looking at individuals that have contributed the most amounts to the 
Presidential Campaigns we see that most of them have top-level positions i.e. 
CEO, President, Founder, Chairman, Owner, etc. Let's also look at the 
organizations they work for.

```{r echo=FALSE, message=FALSE, warning=FALSE}
print("Organizations Top Paying Individuals work for:")
summary(high_payers$contbr_employer)[summary(high_payers$contbr_employer)>0]
```

Here, we see that most of these individuals are either self employed, retired
or have chosen not to reveal their organization (thus, in the category 
"INFORMATION REQUESTED").

Now, lets look at the financial contributions made to candidates of the 
Democratic Party,

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(y=contb_receipt_amt, x=cand_nm), 
       data = subset(Data, 
                     (pol_party=='Democratic Party')&(contb_receipt_amt>0)))+
  geom_boxplot() +
  coord_cartesian(ylim = c(0,1000))+
  xlab('Candidate Name')+
  ylab('Amount')
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
Democratic_Party <- (subset(Data, 
               (pol_party=='Democratic Party')&(contb_receipt_amt>0)))
Democratic_Party$cand_nm <- droplevels(Democratic_Party$cand_nm)

by(Democratic_Party$contb_receipt_amt, Democratic_Party$cand_nm, summary)
```

It is interesting to note that the candidates  who got the most number of 
donations (from univariate analysis) have the lowest median and mean among
all the Democratic Party Candidates. Let's see if this trend persists for
the Republican Party also.

```{r echo=FALSE, message=FALSE, warning=FALSE}
Republican_Party <- (subset(Data, 
               (pol_party=='Republican Party')&(contb_receipt_amt>0)))

#As there are a lot of republican candidates, we will be only plotting those
#with number of donations greater than 3000
top_republicans <- names(summary(Republican_Party$cand_nm)
                         [(summary(Republican_Party$cand_nm)>3000)])

Republican_Party <- subset(Republican_Party,cand_nm %in% top_republicans)
Republican_Party$cand_nm <- droplevels(Republican_Party$cand_nm)

ggplot(aes(y=contb_receipt_amt, x=cand_nm), data = Republican_Party)+
  geom_boxplot() +
  coord_cartesian(ylim = c(0,1000))+
  xlab('Candidate Name')+
  ylab('Amount')+
  theme(axis.text.x = element_text(angle=90, hjust=1)) 
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
by(Republican_Party$contb_receipt_amt, Republican_Party$cand_nm, summary)
```

We can again see that for candidates who received large number of donations,
the median and mean for the amount received is lower compared to the other
candidates.However, the values for the mean and the median were higher for top
Republican candidates than for top Democratic Candidates which may be because 
the number of overall contributions was lower for the Republican party.

Now, that we have looked at each donation, we will look at the total amount
collected by each candidate through these contributions.

```{r echo=FALSE, warning=FALSE, message=FALSE}
dotchart(by(Data$contb_receipt_amt,Data$cand_nm,sum),
         xlab = 'Total Amount ($)',
         main= 'Total Amount Received by Each Candidate',
         lcolor = 'Blue',
         color = 'Black')
```

Here, we can see that the two candidates who received the maximum number of 
donations also, received the highest amounts. However, the total recieved was 
much greater for one of those candidates than the other.

Thus, lets look at the correlation between the number of donations made and 
the total amount generated for each candidate.

```{r echo=FALSE, warning=FALSE, message=FALSE}
cor.test(by(Data$contb_receipt_amt,Data$cand_nm,sum), table(Data$cand_nm))
```

There appears to be a very high correlation between the total amount generated
and the number of donations as shown by the correlation coefficient.

Let's now look at the amount received by candidates on the dates when maximum
donations were made (already calculated in Univariate Section)

```{r echo=FALSE, message=FALSE, warning=FALSE}
top_date_values <- subset(Data, contb_receipt_dt %in% top_dates)
top_date_values$contb_receipt_dt <- 
  droplevels(top_date_values$contb_receipt_dt)

ggplot(aes(x=contb_receipt_dt, y=contb_receipt_amt), 
       data = subset(top_date_values, contb_receipt_amt>0))+
  geom_bar(stat='identity')+
  xlab('Dates')+
  ylab('Amount Contributed ($)')+
  scale_y_continuous(labels = scales::unit_format("Million",1e-6))+
  theme(axis.text.x = element_text(angle=90, hjust=1)) 
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
by(top_date_values$contb_receipt_amt, top_date_values$contb_receipt_dt, sum)
```

From the graph and the values above we find that the largest amount was donated
on 26th September 2016, around one month before the General Election.

Lets also look at the correlation between the number of donations made each day
and the total amount donated that day.

```{r echo=FALSE, message=FALSE, warning=FALSE}
cor.test(by(Data$contb_receipt_amt,
            Data$contb_receipt_dt,sum),
         table(Data$contb_receipt_dt))
```

We can see that the correlation is quite strong as shown by the Pearson's 
Correlation coefficient.

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset

Multiple analysis were performed which included studying the relation between 
gender and amount donated, identifying occupations and employers who have 
contributed the largest single amounts to the campaigns, the variation in 
amounts donated depending on the Candidate among each major political party and
looking at different correlations between various variables.

It was interesting to see that female candidates, had a higher number of 
'large' donations compared to their male counter parts. This was mainly due to
contributions made toward one candidate ("Clinton, Hillary Rodham"). It was 
also seen to no surprise that the large contributions came from indivduals who 
had "high ranking" occupations such as CEO, founder, president, etc. An 
analysis of the relation between total contributions made and the total amount 
donated to a particular candidate showed a very strong correlation exhibited 
by the Pearson Correlation Coefficient.

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?
As the main variable being analysed is "contb_receipt_amt" i.e. the amount 
contributed for each entry, all relations are generated with respect to this 
variable. 

We saw that females generated more money than their male counterparts and
having higher contributions would increase your chances of generating greater 
money. Also, Pearson's Correlation coefficient suggested that greater the 
number of donations made in a day, greater is the amount generated.

### What was the strongest relationship you found?
The strongest relation established was the one between the total number of 
donations made to a candidate and the total amount generated by that candidate.

# Multivariate Plots Section

Let's first look at the candidate who received the most donation amount and
analyze which city contributed the most to this candidate.

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots}
hillary_clinton <- subset(Data, cand_nm == 'Clinton, Hillary Rodham')
hillary_clinton <- subset(hillary_clinton, contb_receipt_amt>0)

hc_city_sums <- aggregate(hillary_clinton$contb_receipt_amt,
                          by=list(hillary_clinton$contbr_city), FUN=sum)

hc_city_sums <- subset(hc_city_sums, x>750000)
hc_city_sums[xx] <- geocode(as.character(hc_city_sums$Group.1))
hc_city_sums <- droplevels(hc_city_sums)

Cal_map+
  geom_point(aes(x=lon, y=lat, size = x/(1e6)), 
             data=subset(hc_city_sums, x>2400000), alpha=0.2)+
  scale_size_area(name="Donation in million $")+
  geom_text(aes(x=lon, y=lat, label=Group.1), hjust="inward", 
            data=subset(hc_city_sums, x>2400000))+
  ggtitle("Cities with most contributions for the Top Receiving Candidate")

```

This is very similar to the plot we got in the univariate section when we 
examined the total contributions made by each city but here, 'Palo Alto' 
replaces 'San Diego' as people living here donated more to the top candidate. 
Let's also look at the individual amounts that were donated to the top receiving
candidate in the top donating cities.

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(aes(x=contbr_city, y=contb_receipt_amt), 
       data=subset(hillary_clinton, contbr_city %in%large_contrib_cities))+
  geom_point(alpha =0.05, position=position_jitter())+
  scale_y_continuous(breaks=seq(0,6500,1000))+
  xlab('City Name')+
  ylab('Amount ($)')+
  theme(axis.text.x = element_text(angle=90, hjust=1)) 
```

We can see that most of the donations made were $1000 or less, but for the 
cities of San Francisco and Los Angeles, the amounts donated greater than $1000
were higher compared to the other cities, this contibuted to them donating the 
most money.

Lets look at the variation in the amount donated per person depending on the 
employer for the top 5 candidates.

```{r message=FALSE, warning=FALSE}
#We will be making two plots, which will make data visualization easier.

#Employers listed on first graph
gp_one <- c("NONE", "HOMEMAKER", "NOT EMPLOYED", "SELF EMPLOYED", "RETIRED")

#Employers listed on second graph
gp_two <- c("APPLE", "GOOGLE", "UCLA", "STATE OF CALIFORNIA", 
            "STANFORD UNIVERSITY")

#Candidates receiving most donations, as seen earlier
cand_name_emp <- c("Clinton, Hillary Rodham", "Sanders, Bernard", 
                   "Trump, Donald J.","Cruz, Rafael Edward 'Ted'", 
                   "Rubio, Marco")

#Function to get the total number of donations per candidate and amount given 
#per donation for each employer

get_data <- function(employers, candidates, dataframe){
  cand_employer_data <- ddply(dataframe, c("contbr_employer","cand_nm"),
                              c("nrow"))
  
  gp_data <- subset(cand_employer_data, contbr_employer %in% employers)
  gp_data <- subset(gp_data, cand_nm %in% candidates)
  
  sums <- aggregate(Data$contb_receipt_amt,by=list(Data$cand_nm,
                                                 Data$contbr_employer),FUN=sum)
  sums <- subset(sums, (Group.1 %in% candidates)&(Group.2 %in% employers))
  
  gp_data$per_person <- sums$x/gp_data$nrow
  return (gp_data)
}

gp_one_data <- get_data(gp_one,cand_name_emp, Data)
gp_two_data <- get_data(gp_two,cand_name_emp, Data)

```

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(y=per_person, x=contbr_employer, fill = cand_nm),
       data=gp_one_data)+
  geom_col(position="dodge")+
  scale_color_brewer(palette = "Dark2")+
  labs(fill="Candidate Name")+
  ylab("Contribution per person")+
  xlab("Employer Name")+
  ggtitle("Variation in Amount contributed per person depending on Employer")
  
```


```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(aes(y=per_person, x=contbr_employer, fill = cand_nm),
       data=gp_two_data)+
  geom_col(position="dodge")+
  scale_color_brewer(palette = "Dark2")+
  ylab("Contribution per person")+
  xlab("Employer Name")+
  labs(fill="Candidate Name")+
  ggtitle("Variation in Amount contributed per person depending on Employer")+
  theme(axis.text.x = element_text(angle=90, hjust=1)) 

```

We can see from the analysis, that "Rubio, Marco" received the greatest 
Contribution per person for most of the cases. In a lot of the cases minimum 
contribution per person was received by "Clinton, Hillary Rodham" who we have 
seen received the most donations and the greatest amount in previous analysis. 
Thus, this greater contribution per person can be because of lesser number of
contributions made to "Rubio, Marco". Thus, we look at total contributions made
by individuals working for each of these employers to the candidates. To plot
these values we use a log scale on the y axis,

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(aes(y=per_person*nrow, x=contbr_employer, fill = cand_nm),
       data=gp_one_data)+
  geom_col(position="dodge")+
  scale_y_log10()+
  scale_color_brewer(palette = "Dark2")+
  ylab("Total Contribution ($)")+
  xlab("Employer Name")+
  labs(fill="Candidate Name")+
  ggtitle("Variation in Total Amount contributed by individuls working 
                                for an Employer")

```

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(y=per_person*nrow, x=contbr_employer, fill = cand_nm),
       data=gp_two_data)+
  geom_col(position="dodge")+
  scale_y_log10()+
  scale_color_brewer(palette = "Dark2")+
  ylab("Total Contribution ($)")+
  xlab("Employer Name")+
  labs(fill="Candidate Name")+
  ggtitle("Variation in Total Amount contributed by individuls working 
                                for an Employer")+
  theme(axis.text.x = element_text(angle=90, hjust=1)) 
```

The two graphs give very different results. The second one confirms the 
hypothesis stated that "Clinton, Hillary Rodham" got the most amount of 
donations because of her larger number of donations, but in the first plot
we see that "Sanders, Bernard", "Rubio, Marco" and "Trump, Donald J." 
received a total greater contribution than "Clinton, Hillary Rodham" for some 
values.

For these values, lets look at the total contributions made and the average
contribution per person,

```{r echo=FALSE, message=FALSE, warning=FALSE}
vals_interest <- subset(gp_one_data,contbr_employer %in% 
                          c("HOMEMAKER", "NOT EMPLOYED",
                                          "NONE", "RETIRED"))
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
subset(vals_interest,contbr_employer %in% "NOT EMPLOYED")
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
subset(vals_interest,contbr_employer %in% "HOMEMAKER")
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
subset(vals_interest,contbr_employer %in% "NONE")
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
subset(vals_interest,contbr_employer %in% "RETIRED")
```

We see that, "Sanders, Bernard" received much greater contributions than anyone
else in the categories with employer names, "None" & "Not Employed" which 
combined with a good contribution per person lead to him having the most total
donations there. 

Within the employee category, "Homemaker", "Trump, Donald J." 
& "Cruz, Rafael Edward 'Ted'" received almost the same total amount of 
contributions despite the fact that "Cruz, Rafael Edward 'Ted'" received almost
200 more donations.

For the employee category, "Retired", we find that although "Clinton, Hillary 
Rodham" received more number of donations but, "Trump, Donald J." received the
greater total amount because of greater amount donated per person.

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?
Two different analysis were performed here, first we looked at the sources of
donations for the top receiving candidate and found that two cities are major
contributors to this amount.

The other analysis examined the possiblity of a relation between employers
of the donor and the candidate receiving the donation. 

### Were there any interesting or surprising interactions between features?
It was interesting to see that donors who work for certain employers deviated 
from the consensus top receiving candidate and having greatest number of 
contributions from a particular group is not a guarantee that the total 
contribution received from that group will be the greatest. Also, if popularity
is measured by the number of donations received, we have identified certain 
groups (Employers under the categories, "Not Employed", "None", "Homemaker" & 
"Retired") which have different preferences from the overall consensus.

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One}
Cal_map+
  geom_point(aes(x=lon, y=lat, size = nrow), data=city_vals, alpha=0.25)+
  scale_size_area(name="No. of Donations")+
  geom_text(aes(x=lon, y=lat, label=contbr_city), hjust="inward", 
            data=subset(city_vals, nrow>35000))+
  ggtitle('Cities with Highest Contributions')+
  xlab('Longitude')+
  ylab('Latitude')
```

### Description One
This plot looks at the cities from where the maximum donations were received,
it can be seen clearly that most of these cities are located along the edges/
coast of the region. This implies that these locations might be densely 
populated and it makes sense to spend more time and resources campaigning in 
these cities in the future.

### Plot Two
```{r echo=FALSE, Plot_Two}
ggplot(aes(y=contb_receipt_amt, x=cand_nm), 
       data = subset(Data, 
                     (pol_party=='Democratic Party')&(contb_receipt_amt>0)))+
  geom_boxplot() +
  coord_cartesian(ylim = c(0,1000))+
  xlab('Candidate Name')+
  ylab('Amount ($)')+
  ggtitle('Variations in amount received by each candidate')
```

### Description Two
From the above plot, it can be seen that the candidates who received the most
number (and amount) of donations have a lower median and quartile values, with 
a lot of outliers as most of these donations are smaller amounts which drives 
the median and quartile values low. This trend is consistent for candidates of 
both the major political parties.

### Plot Three
```{r echo=FALSE, Plot_Three}
ggplot(aes(y=per_person*nrow, x=contbr_employer, fill = cand_nm),
       data=gp_one_data)+
  geom_col(position="dodge")+
  scale_y_log10()+
  scale_color_brewer(palette = "Dark2")+
  ylab("Total Contribution ($)")+
  xlab("Employer Name")+
  labs(fill="Candidate Name")+
  ggtitle("Variation in Total Amount contributed by individuls working 
                                for an Employer")+
  theme(axis.text.x = element_text(angle=90, hjust=1)) 
```

### Description Three
From the above graph we see that individuals working for certain employers have
preferences different from the state consensus (if popularity is taken as the 
number of donations received by a candidate). We see that individuals working 
for 'None', 'Retired', 'Not Employed' and 'Homemaker' prefer different 
candidates to the one most popular in the state ('Clinton, Hillary Rodham'). 
This can serve as an important factor in formulating proposed policies and 
targetting new donors which all contribute to being successful in an election.

------

# Reflection

This dataset contained information about the donations received by candidates
for the 2016 U.S. Presidential Election from the state of California. 

As, there was really only one continous variable in amount donated, analysis 
revolved around its variation with different groups like political party, 
candidate name, city of the donor , employer information & occupation of the 
donor. It was interesting to note that the analysis kept pointing towards a 
single main leader within the candidates (Clinton, Hillary Rodham) which 
received the most number and the highest amount of donations.

Variations in the amount donated to candidates revealed that more the number of
donations, there is a greater likelihood that the median donation received is
lesser. We were also able to identify certain groups that have preferences 
which are different from the overall trend. This is important for targetting 
new 'vote banks' and generating more funds for the campaign.

This analysis could have been more detailed if, some extra information like 
age of donors & donor genders was available as candidate preferences vary a lot
with age and donor gender might have a bearing on the candidate of choice.
The conclusions discussed in the analysis are based on certain assumptions like
popularity is a measured by number of donations received by a candidate. Thus,
care should be taken while drawing conclusions.

This data could be analysed against the data for the Presidential Election held
in 2012 to look at variations in contributions by cities, the differences in 
dates when these contributions were made and looking at the variation in funds
generated between the two major political parties.